---
title: "final project Winning Formula: An Analysis of F1 Race Success (2010-2024)"
author: "Andrew Fouty, Zoe Zhang"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# load library
library(tidyverse)
library(car)
library(boot)
library(ggplot2)
library(dplyr)
library(corrplot)
library(leaps)
library(MASS)
```

# 1. Introduction

Questions:

Is qualifying position a good predictor of race results? 
are the big teams (Red Bull, Ferrari, McLauren, Mercedes) faster than the rest of the field.


season champion: 
1) get raceid from f1_races, max(round) by `year`
2) get driverid from f1_driver_standings, filter on raceid from step 1 and position =1


```{r}



```


# 2. Data cleaning and preparation

The data comes from [Formula 1 World Championship (1950 - 2024)] (https://www.kaggle.com/datasets/rohanrao/formula-1-world-championship-1950-2020).
The dataset consists of all information on the Formula 1 races, drivers, constructors, qualifying, circuits, lap times, pit stops, championships from 1950 till the latest 2024 season.

Looking at races since year 2010 (modern era)
note: this Kaggle/Ergast dataset uses the string ("\N") for missing data. we would need to handle this specifically, making sure numeric columns are numeric as they're supposed to


```{r}
# Load datasets
base_url <- "https://raw.githubusercontent.com/zoe-zg/SDS5300-final-project/main/data/"

f1_results <- read.csv(paste0(base_url, "results.csv"), na.strings = "\\N")
f1_constructor_results <- read.csv(paste0(base_url, "constructor_results.csv"), na.strings = "\\N")
f1_races <- read.csv(paste0(base_url, "races.csv"), na.strings = "\\N")
f1_drivers <- read.csv(paste0(base_url, "drivers.csv"), na.strings = "\\N")
f1_constructors <- read.csv(paste0(base_url, "constructors.csv"), na.strings = "\\N")
f1_driver_standings <- read.csv(paste0(base_url, "driver_standings.csv"), na.strings = "\\N")
f1_constructor_standings <- read.csv(paste0(base_url, "constructor_standings.csv"), na.strings = "\\N")
f1_quali <- read.csv(paste0(base_url, "qualifying.csv"), na.strings = "\\N")
f1_lap_times <- read.csv(paste0(base_url, "lap_times.csv"), na.strings = "\\N")
f1_pit_stops <- read.csv(paste0(base_url, "pit_stops.csv"), na.strings = "\\N")
# f1_status <- read.csv(paste0(base_url, "status.csv"), na.strings = "\\N")
# f1_seasons <- read.csv(paste0(base_url, "seasons.csv"), na.strings = "\\N")

f1_pit_summary <- f1_pit_stops %>% 
  group_by(raceId, driverId) %>% 
  summarize(duration = mean(milliseconds)/1000)

# Merge datasets
f1_data <- f1_results %>%  
  left_join(f1_races, by = "raceId") %>%
  left_join(f1_drivers, by = "driverId") %>%
  left_join(f1_constructors, by = "constructorId") %>% 
  left_join(f1_pit_summary, by = c("raceId", "driverId")) %>% 
  rename(driver_nationality = nationality.x, constructor_nationality = nationality.y,
         race_name = name.x, constructor_name = name.y) %>% 
  dplyr::select(raceId, year, round, driverId, circuitId, race_name, grid, positionOrder, points, 
         driverRef, constructorRef, driver_nationality, constructor_nationality,
         statusId, laps, fastestLapSpeed, fastestLapTime, fastestLap, dob, milliseconds, round, duration) 

# f1_driver_standings <- f1_driver_standings %>% 
#   left_join(f1_drivers, by = "driverId")

```


```{r}

# Text Character Replacement / Cleaning
f1_clean <- f1_data %>% 
  filter(year >= 2010) %>%
  mutate(driverRef = as.factor(driverRef),
         constructorRef = as.factor(constructorRef),
         # create a binary variable for logistic regression
         # did the driver score points (top 10 finish)
         is_points = ifelse(positionOrder <= 10, 1, 0),
         is_points = as.factor(is_points), 
         # did the driver finish on the podium (top 3 finish)
         is_podium = ifelse(positionOrder <=3, 1, 0),
         is_podium = as.factor(is_podium),
         fastestLapSpeed = as.numeric(as.character(fastestLapSpeed)),
         age = year - as.numeric(substr(dob, 1, 4)),
         positionOrder = positionOrder,
         fastestLapTime = as.numeric(ms(f1_clean$fastestLapTime))
         ) %>% 
  # missing values
  drop_na(grid, positionOrder)
  

# dim(f1_clean)
# str(f1_clean)
# head(f1_clean)
# unique(f1_data$constructorRef)
```

# 3. Explorative Analysis

- Histogram 
- Boxplot 
- Scatterplot 
- Q-Q plot 




Identify and visualize the dominance of top teams in the modern era

```{r}
# top 10 activity teams
top_teams <- f1_clean %>%
  count(constructorRef, sort = TRUE) %>%
  top_n(10) %>%
  pull(constructorRef)

par(mar = c(8,4,4,2))
boxplot(positionOrder ~ constructorRef, 
        data = f1_clean %>% filter(constructorRef %in% top_10_teams),
        main = "Finishing Position by Team (2010-2020)",
        ylab = "Finishing Position",
        col = "lightblue",
        las = 2) 

f1_clean %>%
  filter(constructorRef %in% top_teams) %>% 
  mutate(constructorRef = reorder(constructorRef, positionOrder, median)) %>% 
  ggplot(aes(x = constructorRef, y = positionOrder, fill = constructorRef)) +
  geom_boxplot(ooutlier.alpha = 0.3) +
  labs(title = "Finishing Position Distribution by Team (2010-2024)",
       subtitle = "Ordered by Median Finishing Position (Lower is Better)",
       x = "Team",
       y = "Finishing Position") +
  theme_minimal() +
  coord_flip() 

```

```{r}



```


##### histogram and normal quantile
```{r}

ggplot(f1_clean, aes(x = fastestLapSpeed)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Fastest Lap Speeds", x = "Speed (km/h)", y = "Count") +
  theme_minimal()


ggplot(f1_clean, aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Driver's Age", x = "", y = "") +
  theme_minimal()

ggplot(f1_clean, aes(x = fastestLapTime)) +
  geom_histogram(binwidth = , fill = "steelblue", color = "white") +
  labs(title = "", x = "", y = "") +
  theme_minimal()


```




# 4. Basic Tests

question: does driver's performance vary across consstructors, possibly Alonso, Renault and Ferrari
-	median starting position vs is_podium
-	driver’s mean fastestlaptime, at different teams, e.g. Alonso

##### t-test, bootstrap, permutation

```{r}

```


```{r}


```


##### correlation
fastest lap time vs driver's age


```{r}

# cor_est <- cor(f1_clean$grid, f1_clean$positionOrder)
# print(paste("Pearson Correlation:", round(cor_est, 3)))

# f1_clean_sub %>%
#   filter(driverRef %in% top_teams)

plot(jitter(grid, factor = 2) ~ jitter(positionOrder, factor = 2), data = f1_clean_alonso, main = "", 
     xlab = "", ylab = "", pch = 19, col = "blue")
abline(lm(grid ~ positionOrder, data = f1_clean_alonso), col = "red")


lm_alonso <- lm(grid ~ positionOrder, data = f1_clean_alonso)
summary(lm_alonso)

r <- rstudent(lm_alonso)
f <- fitted(lm_alonso)

plot(r ~ f)

source("https://raw.githubusercontent.com/jreuning/sds230_files/refs/heads/main/regJDRS.txt")
myResPlots(lm_alonso, label = "")

# f1_clean_alonso <- f1_clean[f1_clean$driverRef == "alonso", ]
# 
# plot(jitter(fastestLapSpeed) ~ jitter(age), data = f1_clean_hamiltom, main = "", 
#      xlab = "", ylab = "", pch = 19, col = "blue")
# abline(lm(fastestLapSpeed ~ age, data = f1_clean_hamiltom), col = "red")
# 
# 
# plot(jitter(grid) ~ jitter(age), data = f1_clean_hamiltom, main = "", 
#      xlab = "", ylab = "", pch = 19, col = "blue")
# abline(lm(grid ~ age, data = f1_clean_hamiltom), col = "red")

```




```{r}

# --- Comparison of Top Drivers ---
top_drivers <- c("hamilton", "vettel", "alonso", "max_verstappen", "norris", "piastri", "leclerc", "russell", "sainz", "michael_schumacher")

f1_clean %>%
  filter(driverRef %in% top_drivers) %>%
  ggplot(aes(x = grid, y = positionOrder, color = driverRef)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Qualifying vs. Race Result: Top Drivers Comparison",
       x = "Starting Grid Position",
       y = "Finishing Position") +
  theme_minimal() +
  facet_wrap(~driverRef) # Create separate plot for each driver


# dataset
f1_driver_data <- f1_clean %>% 
  filter(driverRef %in% top_drivers) %>%
  left_join(f1_driver_standings %>% select(raceId, driverId, position), 
            by = c("raceId", "driverId"), suffix = c("", "_standing")) %>%
  mutate(fastestLapSpeed = as.numeric(as.character(fastestLapSpeed)))
            

cor_data <- f1_driver_data %>%
  select(grid, positionOrder, fastestLapSpeed, points) %>%
  drop_na()

  
  corrplot(cor(cor_data), method = "circle", type = "upper", 
         title = "Correlation Matrix: Top Drivers (2010-2020)", 
         mar = c(0,0,2,0), 
         addCoef.col = "black") # Add numbers
```



# 5. Multiple Linear Regression (Stepwise)


```{r}

# Prepare data for model (filter to top teams to handle categorical levels)
model_data <- f1_clean %>%
  filter(constructorRef %in% top_teams) %>%
  select(positionOrder, grid, constructorRef, year) %>%
  drop_na()

# Full Model
full_lm <- lm(positionOrder ~ grid + constructorRef + year, data = model_data)

# Backward Stepwise Selection
step_lm <- step(full_lm, direction = "backward", trace = 0) # trace=0 hides output
summary(step_lm)

```

```{r}
par(mfrow=c(2,2))
plot(step_lm)

```








Top 15 fastest circuits
```{r}
top_15_names <- f1_clean %>%
  group_by(name) %>%
  summarise(mean_speed = mean(fastestLapSpeed)) %>%
  slice_max(mean_speed, n = 15) %>%
  pull(name)

top_15_data <- circuit_speeds %>% filter(name %in% top_15_names)

ggplot(top_15_data, aes(x = reorder(name, fastestLapSpeed), y = fastestLapSpeed, fill = name)) +
  # The Violin shows the distribution shape (density)
  geom_violin(trim = FALSE, alpha = 0.6) +

  # Adding a narrow boxplot inside gives you the median/quartiles (Best of both worlds)
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  
  coord_flip() + # Horizontal is easier to read
  
  labs(title = "Top 15 Fastest F1 Circuits (Violin Plot)",
       subtitle = "Shape represents distribution of lap speeds. White bar is the interquartile range.",
       x = NULL,
       y = "Fastest Lap Speed (km/h)") +
  theme_minimal() +
  theme(legend.position = "none")
  
# distribution of speeds of all races
ggplot(circuit_speeds, aes(x = fastestLapSpeed)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white", alpha = 0.7) +
  labs(title = "Distribution of Fastest Lap Speeds Across All Races",
       x = "Speed (km/h)",
       y = "Count") +
  theme_minimal()
```



```{r}
# Top 5 Teams only (to keep categorical variables manageable)

f1_final <- f1_clean %>%
  filter(constructorRef %in% top_teams) %>%
  droplevels()

# --- 2. Graphics ---
par(mfrow=c(2,2)) # Set up 2x2 plotting grid

# 1. Scatterplot: Grid vs Finish
plot(positionOrder ~ grid, data = f1_final, main = "Grid vs. Finish Position", 
     xlab = "Grid Position", ylab = "Finish Position", pch = 19, col = "blue")
abline(lm(positionOrder ~ grid, data = f1_final), col = "red")

# 2. Boxplot: Finish by Team
boxplot(positionOrder ~ constructorRef, data = f1_final, main = "Finish by Team", 
        las = 2, col = "lightblue")

# 3. Histogram: Distribution of Grid Positions
hist(f1_final$grid, main = "Distribution of Grid Positions", 
     xlab = "Grid Position", col = "lightgreen", border = "white")

# 4. Normal Quantile Plot (on residuals of a simple model)
simple_model <- lm(positionOrder ~ grid, data = f1_final)
qqnorm(residuals(simple_model), main = "Normal Q-Q Plot of Residuals")
qqline(residuals(simple_model), col = "red")

par(mfrow=c(1,1)) # Reset plotting grid

# --- 3. Basic Tests ---
# 1. T-test: Compare average finish position of Mercedes vs. Ferrari
mercedes_pos <- f1_final$positionOrder[f1_final$constructorRef == "mercedes"]
ferrari_pos <- f1_final$positionOrder[f1_final$constructorRef == "ferrari"]

t_test_result <- t.test(mercedes_pos, ferrari_pos)
print(t_test_result)
# Interpretation: If p-value < 0.05, there is a significant difference in mean finish position.

# 2. Correlation
cor_val <- cor(f1_final$grid, f1_final$positionOrder)
cat("Correlation between Grid and Finish:", cor_val, "\n")

# 3. Bootstrap CI for Correlation
corr_fun <- function(data, indices) {
  d <- data[indices, ]
  return(cor(d$grid, d$positionOrder))
}
set.seed(123)
boot_results <- boot(data = f1_final, statistic = corr_fun, R = 1000)
print(boot.ci(boot_results, type = "perc"))

# --- 4. Permutation Test ---
# Hypothesis: Does Lewis Hamilton finish significantly better (lower position) than the average of all other drivers?
# Test Statistic: Difference in means (Others - Hamilton). Positive value means Hamilton is better (lower score).

obs_diff <- mean(f1_final$positionOrder[f1_final$driverRef != "hamilton"]) - 
            mean(f1_final$positionOrder[f1_final$driverRef == "hamilton"])

cat("Observed Difference (Others - Hamilton):", obs_diff, "\n")

```


 Permutation test
```{r}
# Permutation Loop
set.seed(123)
n_perms <- 1000
perm_diffs <- numeric(n_perms)
all_positions <- f1_final$positionOrder

for (i in 1:n_perms) {
  # Shuffle the positions
  shuffled_pos <- sample(all_positions)
  
  # Assign shuffled positions to groups based on original indices
  # (We keep the group sizes fixed: Hamilton vs Others)
  hamilton_indices <- which(f1_final$driverRef == "hamilton")
  
  mean_hamilton_perm <- mean(shuffled_pos[hamilton_indices])
  mean_others_perm <- mean(shuffled_pos[-hamilton_indices])
  
  perm_diffs[i] <- mean_others_perm - mean_hamilton_perm
}

# P-value: Proportion of permutations where the difference is as extreme or more extreme than observed
p_value <- mean(perm_diffs >= obs_diff)
cat("Permutation Test P-value:", p_value, "\n")
hist(perm_diffs, main = "Permutation Distribution", xlab = "Difference in Means", col = "gray")
abline(v = obs_diff, col = "red", lwd = 2)

```


Multiple Regression 
```{r}
# --- 5. Multiple Regression ---
# Using Stepwise Regression (Backward)
# Model: Predict Finish Position based on Grid, Team, and Year
full_model <- lm(positionOrder ~ grid + constructorRef + year, data = f1_final)
step_model <- step(full_model, direction = "backward", trace = 0) # trace=0 suppresses output

summary(step_model)

# Residual Plots for the final model
par(mfrow=c(2,2))
plot(step_model)
par(mfrow=c(1,1))

# Interpretation:
# - Intercept: Baseline finish position.
# - Grid: For every 1 spot further back on the grid, finish position increases by approx [coefficient].
# - Constructor: Coefficients show difference relative to the reference team.
# - Year: Effect of time on finish position (if significant).

# --- 6. Advanced Technique: Binary Logistic Regression ---
# Predict probability of finishing in the points (Top 10)
log_model <- glm(in_points ~ grid + constructorRef, data = f1_final, family = "binomial")
summary(log_model)

# Interpretation of Odds Ratios
odds_ratios <- exp(coef(log_model))
cat("Odds Ratios:\n")
print(odds_ratios)
# Example: If Grid OR < 1, starting further back decreases odds of points.
# Example: If Mercedes OR > 1, being in Mercedes increases odds of points vs baseline team.
```


# Delete

Permutation test
We test if Max Verstappen's average finishing position is significantly better (lower) than the population mean using a permutation test.

```{r}

# Observed difference
verstappen_mean <- mean(f1_clean$positionOrder[f1_clean$driverRef == "max_verstappen"])
population_mean <- mean(f1_clean$positionOrder)
obs_diff <- population_mean - verstappen_mean # Positive means Verstappen is faster (lower position)

# Permutation loop
set.seed(123)
n_perms <- 2000
perm_diffs <- numeric(n_perms)

for(i in 1:n_perms) {
  # Shuffle the driver labels
  shuffled_drivers <- sample(f1_clean$driverRef)
  
  # Calculate mean for "fake" Hamiltons in this shuffled set
  perm_verstappebn_mean <- mean(f1_clean$positionOrder[shuffled_drivers == "max_verstappen"])
  perm_diffs[i] <- population_mean - perm_verstappebn_mean
}

# Calculate p-value
p_val <- mean(perm_diffs >= obs_diff)
print(paste("Observed Diff:", round(obs_diff, 2)))
print(paste("P-value:", p_val))

```


##### plot
```{r}
par(mfrow=c(2,2)) 

# 1. Histogram: Distribution of Fastest Lap Speeds
hist(f1_clean$fastestLapSpeed, 
     main = "Histogram of Fastest Lap Speeds", 
     xlab = "Speed (km/h)", col = "skyblue", border = "white")

# 2. Boxplot: Starting Grid Position vs. Did they Score Points?
boxplot(grid ~ is_points, data = f1_clean,
        main = "Grid Position vs. Points Finish",
        xlab = "Scored Points (0=No, 1=Yes)", ylab = "Grid Position",
        col = c("lightcoral", "skyblue"))


boxplot(grid ~ driverRef , data = f1_clean,
        main = "Grid Position vs. Points Finish",
        xlab = "Scored Points (0=No, 1=Yes)", ylab = "Grid Position",
        col = c("lightcoral", "skyblue"))

# 3. Scatterplot: Grid Position vs. Finishing Position
plot(f1_clean$grid, f1_clean$positionOrder,
     main = "Scatterplot: Grid vs. Finish",
     xlab = "Starting Grid Position", ylab = "Finishing Position",
     pch = 19, col = rgb(0,0,0,0.2)) 

# 4. Normal Quantile Plot (QQ Plot)
# Checking if Lap Speeds are normally distributed
qqnorm(f1_clean$fastestLapSpeed, main = "QQ Plot: Lap Speeds")
qqline(f1_clean$fastestLapSpeed, col = "red")

```

##### Driver Quali vs Race Result Analysis
Question: how strong does qualifying performance (grid) predict race performance (positionOrder) for specific drivers, and overall? 

Driver 1: Lewis Hamilton
```{r}

driver_data_hamilton <- f1_clean %>% filter(driverRef == "hamilton")
correlation <- cor(driver_data_hamilton$grid, driver_data_hamilton$positionOrder, use = "complete.obs")

ggplot(driver_data_hamilton, aes(x = grid, y = positionOrder)) + 
  geom_point(color = "lightblue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "orange", se = FALSE) +
  labs(
    title = paste("Qualifying vs. Race Result:", toupper("Lewis Hamilton")),
    subtitle = paste("Correlation:", round(correlation, 2)),
    x = "Starting Grid Position", 
    y = "Finishing Position") +
  theme_minimal()
  
```


##### fastest average speed by circuit
Question: which circuits are the fastest? speed distribution

Fastest lap speed across all tracks
```{r}


# fastest lap speed comparison across all tracks
f1_clean %>%
  ggplot(aes(x = race_name, y = fastestLapSpeed, fill = race_name)) +
  geom_boxplot(outlier.size = 0.5, alpha = 0.8) +
  labs(title = "Fastest Lap Speed Comparison across tracks",
       x = NULL,
       y = "Fastest Lap Speed (km/h)") +
  coord_flip() +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.y = element_text(size = 6))
```


##### [explorative] f1dataR package 
```{r}
# install.packages("f1dataR")
# library(f1dataR)

if (!require("remotes")) install.packages("remotes")
remotes::install_github("SCasanova/f1dataR")
library(f1dataR)
```

Telemetry plotting
```{r}
get_driver_name()

f1_laps <- load_laps(season = 2025, round = 15)
f1_driver_telemetry <- load_driver_telemetry(season = "current", round = "last", session = "R", driver = "all", laps = "all")

ham <- load_driver_telemetry(
    season = 2023,
    round = "Bahrain",
    session = "Q",
    driver = "HAM",
    laps = "fastest"
  )
ham <- load_driver_telemetry(2022, 1, "Q", driver = "HAM", laps = "fastest")
per <- load_driver_telemetry(2022, 1, "Q", driver = "PER", laps = "fastest")

telem <- bind_rows(lec, ham, per) %>%
  select(rpm, speed, n_gear, throttle, brake, drs, distance, time, driver_code) %>%
  mutate(drs = ifelse(drs == 12, 1, 0))

drivercolours <- c(
  "LEC" = get_driver_color("LEC", 2022, 1),
  "HAM" = get_driver_color("HAM", 2022, 1),
  "PER" = get_driver_color("PER", 2022, 1)
)
```


