---
title: "final project Winning Formula: An Analysis of F1 Race Success (2010-2024)"
author: "Andrew Fouty, Zoe Zhang"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# load library
library(tidyverse)
library(car)
library(boot)
library(ggplot2)
library(dplyr)
library(corrplot)
library(leaps)
library(MASS)
library(lubridate)
library(PerformanceAnalytics)
```

# 1. Introduction

Questions:

Is qualifying position a good predictor of race results? 
are the big teams (Red Bull, Ferrari, McLauren, Mercedes) faster than the rest of the field.


season champion: 
1) get raceid from f1_races, max(round) by `year`
2) get driverid from f1_driver_standings, filter on raceid from step 1 and position =1


```{r}



```


# 2. Data cleaning and preparation

The data comes from [Formula 1 World Championship (1950 - 2024)] (https://www.kaggle.com/datasets/rohanrao/formula-1-world-championship-1950-2020).
The dataset consists of all information on the Formula 1 races, drivers, constructors, qualifying, circuits, lap times, pit stops, championships from 1950 till the latest 2024 season.



```{r}
# Load datasets
base_url <- "https://raw.githubusercontent.com/zoe-zg/SDS5300-final-project/main/data/"

f1_results <- read.csv(paste0(base_url, "results.csv"), na.strings = "\\N")
f1_constructor_results <- read.csv(paste0(base_url, "constructor_results.csv"), na.strings = "\\N")
f1_races <- read.csv(paste0(base_url, "races.csv"), na.strings = "\\N")
f1_drivers <- read.csv(paste0(base_url, "drivers.csv"), na.strings = "\\N")
f1_constructors <- read.csv(paste0(base_url, "constructors.csv"), na.strings = "\\N")
f1_driver_standings <- read.csv(paste0(base_url, "driver_standings.csv"), na.strings = "\\N")
f1_constructor_standings <- read.csv(paste0(base_url, "constructor_standings.csv"), na.strings = "\\N")
f1_quali <- read.csv(paste0(base_url, "qualifying.csv"), na.strings = "\\N")
f1_lap_times <- read.csv(paste0(base_url, "lap_times.csv"), na.strings = "\\N")
f1_pit_stops <- read.csv(paste0(base_url, "pit_stops.csv"), na.strings = "\\N")
# f1_status <- read.csv(paste0(base_url, "status.csv"), na.strings = "\\N")
# f1_seasons <- read.csv(paste0(base_url, "seasons.csv"), na.strings = "\\N")

f1_pit_summary <- f1_pit_stops %>% 
  group_by(raceId, driverId) %>% 
  summarize(duration = mean(milliseconds)/1000)

# Merge datasets
f1_data <- f1_results %>%  
  left_join(f1_races, by = "raceId") %>%
  left_join(f1_drivers, by = "driverId") %>%
  left_join(f1_constructors, by = "constructorId") %>% 
  left_join(f1_pit_summary, by = c("raceId", "driverId")) %>% 
  rename(driver_nationality = nationality.x, constructor_nationality = nationality.y,
         race_name = name.x, constructor_name = name.y) %>% 
  dplyr::select(raceId, year, round, driverId, circuitId, race_name, grid, positionOrder, points, positionText,
         driverRef, constructorRef, driver_nationality, constructor_nationality,
         statusId, laps, fastestLapSpeed, fastestLapTime, fastestLap, dob, milliseconds, round, duration) 

# f1_driver_standings <- f1_driver_standings %>% 
#   left_join(f1_drivers, by = "driverId")

```


```{r}

# Text Character Replacement / Cleaning
f1_data <- f1_data %>% mutate(fastestLapSeconds = as.numeric(ms(fastestLapTime)))

f1_clean <- f1_data %>% 
  filter(year >= 2010) %>%
  mutate(driverRef = as.factor(driverRef),
         constructorRef = as.factor(constructorRef),
         # create a binary variable for logistic regression
         # did the driver score points (top 10 finish)
         is_points = ifelse(positionOrder <= 10, 1, 0),
         is_points = as.factor(is_points), 
         # did the driver finish on the podium (top 3 finish)
         is_podium = ifelse(positionOrder <=3, 1, 0),
         is_podium = as.factor(is_podium),
         fastestLapSpeed = as.numeric(as.character(fastestLapSpeed)),
         age = year - as.numeric(substr(dob, 1, 4)),
         race_completion = positionText
         ) %>% 
  # missing values
  drop_na(grid, positionOrder)
  



# dim(f1_clean)
# str(f1_clean)
# head(f1_clean)
# unique(f1_data$constructorRef)
```

# 3. Explorative Analysis

- Histogram 
- Boxplot 
- Scatterplot 
- Q-Q plot 

```{r}
# plot theme 

sds5300 <- theme_classic(base_size = 14) +
  theme(text = element_text(family = "serif"),
        axis.text = element_text(color = "black"),
        plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
        panel.grid.major.y = element_line(color = "grey90"),
        panel.grid.major.x = element_line(color = "grey90")
        )

```



Identify and visualize the dominance of top teams in the modern era



```{r}



```

##### boxplot 
- Britain vs Spain drivers race finishing records comparison
- Ferrari vs Red Bull vs McLaren vs Mercedes teams pit stop efficiency comparison


###### 1) drivers' race finishing position

```{r}

# driver counts by country

f1_clean %>% group_by(driver_nationality) %>% 
  summarize(unique_drivers = n_distinct(driverRef)) %>% 
  arrange(desc(unique_drivers))

f1_brit_german <- f1_clean %>% filter(driver_nationality %in% c("British", "German"))

boxplot(positionOrder ~ driver_nationality,
        data = f1_brit_german,
        main = "Race Finishing Position: British vs German Drivers",
        ylab = "Finishing Position",
        xlab = ""
          )

f1_country_most_driver <- f1_clean %>% filter(driver_nationality %in% c("British", "German", "Brazilian", "French", "Spanish"))

boxplot(positionOrder ~ driver_nationality, 
        data = f1_country_most_driver,
        main = "Finishing Positions by Country",
        xlab = "",
        ylab = "Finishing Position",
        las = 1)


boxplot(fastestLapSpeed ~ constructorRef, 
        data = f1_country_most_driver,
        main = "Fastest Lap Speed by Team",
        xlab = "",
        ylab = "Fastest Lap Speed (km/h)",
        las = 2)

```

###### 2) top constructors' pit stop efficiency 

```{r}
boxplot(log(duration) ~ constructorRef, 
        data = f1_clean_pit,
        main = "Pit stop duration by constructor",
        xlab = "",
        ylab = "Pit Stop Duration (seconds)",
        las = 2)
```



##### histogram
- fastest lap speed
- driver's age
- fastest lap time
- mean pit stop time

We filtered out pit stop records exceeding 60 seconds, as these extreme values are most likely outliers caused by race stoppages (e.g. red flags) rather than active competitive pit stops.

```{r}
par(mfrow = c(2, 2))
f1_clean_pit <- f1_clean %>% filter(duration <= 60)

hist(f1_clean$fastestLapSpeed,
     breaks = 20,
     col = "grey40", 
     border = "white",
     main = "Fastest Lap Speeds",
     xlab = "Speed (km/h)",
     las = 1) 

hist(f1_clean$age,
     breaks = 15,
     col = "grey40",
     border = "white",
     main = "Driver's Age",
     xlab = "Age (Years)",
     las = 1)

hist(f1_clean$fastestLapSeconds,
     breaks = 20,
     col = "grey40",
     border = "white",
     main = "Fastest Lap Times",
     xlab = "Time (seconds)",
     las = 1)

hist(f1_clean_pit$duration,
     breaks = 40, # More breaks to see the "two peaks" detail
     col = "grey40",
     border = "white",
     main = "Pit Stop Duration (<60s)",
     xlab = "Duration (seconds)",
     las = 1)

```

##### Q-Q normal quantile plot

```{r}

par(mfrow = c(2, 2))
qqPlot(f1_clean$fastestLapSpeed, 
       main = "QQ Plot: Fastest Lap Speed",
       ylab = "Speed (km/h)",
       col = "red", pch = 19, cex = 0.5)

qqPlot(f1_clean$age, 
       main = "QQ Plot: Driver's Age",
       ylab = "Age (Years",
       col = "red", pch = 19, cex = 0.5)

qqPlot(f1_clean$fastestLapSeconds, 
       main = "QQ Plot: Fastest Lap Time",
       ylab = "Time (Seconds)",
       col = "red", pch = 19, cex = 0.5)

qqPlot(log10(f1_clean_pit$duration), 
       main = "QQ Plot: QQ Plot: Pit Stop Duration (<60s)",
       ylab = "Duration (Seconds)",
       col = "red", pch = 19, cex = 0.5)

```


##### scatter plot
###### 1) Alonso: quali vs. race result
```{r}
f1_alonso <- f1_data[f1_data$driverRef == "alonso", ]

ggplot(f1_alonso, aes( x = grid, y = positionOrder)) +
  geom_jitter(color = "lightblue", alpha = 0.6, size = 4) + 
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_smooth(method = "lm", color = "black", se = FALSE, size = 0.5) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  scale_y_continuous(breaks = seq(0, 20, 5)) +
  labs(title = "Qualifying vs. Race Results: Fernando Alonso",
       subtitle = "Points below the dashed line indicate positions gained during the race",
       x = "Starting Grid Position",
       y = "Finishing Position") +
    theme_linedraw()

```

###### 2) top drivers' quali vs race result
```{r}

# --- Comparison of Top Drivers ---
top_drivers <- c("hamilton", "vettel", "alonso", "max_verstappen", "norris", "piastri", "leclerc", "russell", "sainz", "michael_schumacher")

f1_clean %>%
  filter(driverRef %in% top_drivers) %>%
  ggplot(aes(x = grid, y = positionOrder, color = driverRef)) +
  geom_point(alpha = 0.4) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  # geom_smooth(method = "lm", color = "grey", se = FALSE, size = 0.5) +
  labs(title = "Qualifying vs. Race Result: Top Drivers Comparison",
       x = "Starting Grid Position",
       y = "Finishing Position") +
  theme_linedraw()  +
  facet_wrap(~driverRef) +
  theme(legend.position = "none", 
        strip.text = element_text(size = 12, face = "bold", color = "Black"),
        strip.background = element_rect(fill = "grey90"))

# cor_data <- f1_clean %>%
#   select(grid, positionOrder, fastestLapSpeed, points) %>%
#   drop_na()
# 
#   
# corrplot(cor(cor_data), method = "circle", type = "upper", 
#          title = "Correlation Matrix: Top Drivers (2010-2020)", 
#          mar = c(0,0,2,0), 
#          addCoef.col = "black") # Add numbers
```

###### 3) drivers' fastest lap time at Silverstone by year (2004-2024)

```{r}

f1_britGP <- f1_data %>% filter(race_name == "British Grand Prix", year >= 2004) %>% 
  mutate(decade = floor(year / 10) * 10, decade_label = as.factor(paste0(decade, "s")))

scatterplot(fastestLapSeconds ~ year, 
        data = f1_britGP,
        main = "Fastest Lap Time by Country",
        xlab = "",
        ylab = "Fastest Lap Time (seconds)",
        ylim = c(60, 120),
        las = 1)
```

##### correlation matrix

```{r}
f1_corr <- f1_clean %>% dplyr::select(positionOrder, grid, fastestLapSpeed, year, age, duration)

corrplot(cor(f1_corr), method = "color",
         type = "upper", order = "hclust",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45,
         )

plot(f1_corr, pch = 19, cex = .7, col = 'red', main = "Matrix plot of WB raw data")

chart.Correlation(f1_corr, histogram = TRUE, pch = 19, main = "Matrix plot of WB raw data")
library(GGally)
ggcorr(f1_corr, label = T, size = 4, label_size = 4)
```






# 4. Tests (t-test, bootstrap, permutation)

question: does driver's performance vary across consstructors, possibly Alonso, Renault and Ferrari
-	median starting position vs is_podium
-	driver’s mean fastestlaptime, at different teams, e.g. Alonso

##### case 1: are finishing positions different between British and German drivers

```{r}
ttest_brit_german <- t.test(positionOrder ~ driver_nationality, data = f1_brit_german, conf.level = 0.99)
ttest_brit_german

# p value lower than 0.01, CI doesn't include 0, reject null hypothesis, there is evidence of a difference between British and German drivers
```
```{r}
# To make grading easier, please leave the following line of code in your assignment
set.seed(230) 

# FILL IN REMAINING CODE
n_samp <- 10000

diff_brit_german <- rep(NA, n_samp)

for (i in 1:n_samp) {
  brit <- sample(f1_brit_german$positionOrder[f1_brit_german$driver_nationality == "British"], 
                 sum(iron$Rock == "Sand"), replace = TRUE)
  lime <- sample(iron$logIron[iron$Rock == "Lime"],
                 sum(iron$Rock == "Lime"), replace = TRUE)
  diffLogIron[i] <- mean(lime) - mean(sand)
}


```



```{r}




```


##### correlation

```{r}

# cor_est <- cor(f1_clean$grid, f1_clean$positionOrder)
# print(paste("Pearson Correlation:", round(cor_est, 3)))

# f1_clean_sub %>%
#   filter(driverRef %in% top_teams)

plot(jitter(grid, factor = 2) ~ jitter(positionOrder, factor = 2), data = f1_clean_alonso, main = "", 
     xlab = "", ylab = "", pch = 19, col = "blue")
abline(lm(grid ~ positionOrder, data = f1_clean_alonso), col = "red")


lm_alonso <- lm(grid ~ positionOrder, data = f1_clean_alonso)
summary(lm_alonso)

r <- rstudent(lm_alonso)
f <- fitted(lm_alonso)

plot(r ~ f)

source("https://raw.githubusercontent.com/jreuning/sds230_files/refs/heads/main/regJDRS.txt")
myResPlots(lm_alonso, label = "")


# 
# plot(jitter(fastestLapSpeed) ~ jitter(age), data = f1_clean_hamiltom, main = "", 
#      xlab = "", ylab = "", pch = 19, col = "blue")
# abline(lm(fastestLapSpeed ~ age, data = f1_clean_hamiltom), col = "red")
# 
# 
# plot(jitter(grid) ~ jitter(age), data = f1_clean_hamiltom, main = "", 
#      xlab = "", ylab = "", pch = 19, col = "blue")
# abline(lm(grid ~ age, data = f1_clean_hamiltom), col = "red")

```








# 5. Multiple Linear Regression (Stepwise)


```{r}

# Prepare data for model (filter to top teams to handle categorical levels)
model_data <- f1_clean %>%
  filter(constructorRef %in% top_teams) %>%
  select(positionOrder, grid, constructorRef, year) %>%
  drop_na()

# Full Model
full_lm <- lm(positionOrder ~ grid + constructorRef + year, data = model_data)

# Backward Stepwise Selection
step_lm <- step(full_lm, direction = "backward", trace = 0) # trace=0 hides output
summary(step_lm)

```

```{r}
par(mfrow=c(2,2))
plot(step_lm)

```








Top 15 fastest circuits
```{r}
top_15_names <- f1_clean %>%
  group_by(name) %>%
  summarise(mean_speed = mean(fastestLapSpeed)) %>%
  slice_max(mean_speed, n = 15) %>%
  pull(name)

top_15_data <- circuit_speeds %>% filter(name %in% top_15_names)

ggplot(top_15_data, aes(x = reorder(name, fastestLapSpeed), y = fastestLapSpeed, fill = name)) +
  # The Violin shows the distribution shape (density)
  geom_violin(trim = FALSE, alpha = 0.6) +

  # Adding a narrow boxplot inside gives you the median/quartiles (Best of both worlds)
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  
  coord_flip() + # Horizontal is easier to read
  
  labs(title = "Top 15 Fastest F1 Circuits (Violin Plot)",
       subtitle = "Shape represents distribution of lap speeds. White bar is the interquartile range.",
       x = NULL,
       y = "Fastest Lap Speed (km/h)") +
  theme_bw() +
  theme(legend.position = "none")
  
# distribution of speeds of all races
ggplot(circuit_speeds, aes(x = fastestLapSpeed)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white", alpha = 0.7) +
  labs(title = "Distribution of Fastest Lap Speeds Across All Races",
       x = "Speed (km/h)",
       y = "Count") +
  theme_bw()
```



```{r}
# Top 5 Teams only (to keep categorical variables manageable)

f1_final <- f1_clean %>%
  filter(constructorRef %in% top_teams) %>%
  droplevels()

# --- 2. Graphics ---
par(mfrow=c(2,2)) # Set up 2x2 plotting grid

# 1. Scatterplot: Grid vs Finish
plot(positionOrder ~ grid, data = f1_final, main = "Grid vs. Finish Position", 
     xlab = "Grid Position", ylab = "Finish Position", pch = 19, col = "blue")
abline(lm(positionOrder ~ grid, data = f1_final), col = "red")

# 2. Boxplot: Finish by Team
boxplot(positionOrder ~ constructorRef, data = f1_final, main = "Finish by Team", 
        las = 2, col = "lightblue")

# 3. Histogram: Distribution of Grid Positions
hist(f1_final$grid, main = "Distribution of Grid Positions", 
     xlab = "Grid Position", col = "lightgreen", border = "white")

# 4. Normal Quantile Plot (on residuals of a simple model)
simple_model <- lm(positionOrder ~ grid, data = f1_final)
qqnorm(residuals(simple_model), main = "Normal Q-Q Plot of Residuals")
qqline(residuals(simple_model), col = "red")

par(mfrow=c(1,1)) # Reset plotting grid

# --- 3. Basic Tests ---
# 1. T-test: Compare average finish position of Mercedes vs. Ferrari
mercedes_pos <- f1_final$positionOrder[f1_final$constructorRef == "mercedes"]
ferrari_pos <- f1_final$positionOrder[f1_final$constructorRef == "ferrari"]

t_test_result <- t.test(mercedes_pos, ferrari_pos)
print(t_test_result)
# Interpretation: If p-value < 0.05, there is a significant difference in mean finish position.

# 2. Correlation
cor_val <- cor(f1_final$grid, f1_final$positionOrder)
cat("Correlation between Grid and Finish:", cor_val, "\n")

# 3. Bootstrap CI for Correlation
corr_fun <- function(data, indices) {
  d <- data[indices, ]
  return(cor(d$grid, d$positionOrder))
}
set.seed(123)
boot_results <- boot(data = f1_final, statistic = corr_fun, R = 1000)
print(boot.ci(boot_results, type = "perc"))

# --- 4. Permutation Test ---
# Hypothesis: Does Lewis Hamilton finish significantly better (lower position) than the average of all other drivers?
# Test Statistic: Difference in means (Others - Hamilton). Positive value means Hamilton is better (lower score).

obs_diff <- mean(f1_final$positionOrder[f1_final$driverRef != "hamilton"]) - 
            mean(f1_final$positionOrder[f1_final$driverRef == "hamilton"])

cat("Observed Difference (Others - Hamilton):", obs_diff, "\n")

```


 Permutation test
```{r}
# Permutation Loop
set.seed(123)
n_perms <- 1000
perm_diffs <- numeric(n_perms)
all_positions <- f1_final$positionOrder

for (i in 1:n_perms) {
  # Shuffle the positions
  shuffled_pos <- sample(all_positions)
  
  # Assign shuffled positions to groups based on original indices
  # (We keep the group sizes fixed: Hamilton vs Others)
  hamilton_indices <- which(f1_final$driverRef == "hamilton")
  
  mean_hamilton_perm <- mean(shuffled_pos[hamilton_indices])
  mean_others_perm <- mean(shuffled_pos[-hamilton_indices])
  
  perm_diffs[i] <- mean_others_perm - mean_hamilton_perm
}

# P-value: Proportion of permutations where the difference is as extreme or more extreme than observed
p_value <- mean(perm_diffs >= obs_diff)
cat("Permutation Test P-value:", p_value, "\n")
hist(perm_diffs, main = "Permutation Distribution", xlab = "Difference in Means", col = "gray")
abline(v = obs_diff, col = "red", lwd = 2)

```


Multiple Regression 
```{r}
# --- 5. Multiple Regression ---
# Using Stepwise Regression (Backward)
# Model: Predict Finish Position based on Grid, Team, and Year
full_model <- lm(positionOrder ~ grid + constructorRef + year, data = f1_final)
step_model <- step(full_model, direction = "backward", trace = 0) # trace=0 suppresses output

summary(step_model)

# Residual Plots for the final model
par(mfrow=c(2,2))
plot(step_model)
par(mfrow=c(1,1))

# Interpretation:
# - Intercept: Baseline finish position.
# - Grid: For every 1 spot further back on the grid, finish position increases by approx [coefficient].
# - Constructor: Coefficients show difference relative to the reference team.
# - Year: Effect of time on finish position (if significant).

# --- 6. Advanced Technique: Binary Logistic Regression ---
# Predict probability of finishing in the points (Top 10)
log_model <- glm(in_points ~ grid + constructorRef, data = f1_final, family = "binomial")
summary(log_model)

# Interpretation of Odds Ratios
odds_ratios <- exp(coef(log_model))
cat("Odds Ratios:\n")
print(odds_ratios)
# Example: If Grid OR < 1, starting further back decreases odds of points.
# Example: If Mercedes OR > 1, being in Mercedes increases odds of points vs baseline team.
```


# Delete

Permutation test
We test if Max Verstappen's average finishing position is significantly better (lower) than the population mean using a permutation test.

```{r}

# Observed difference
verstappen_mean <- mean(f1_clean$positionOrder[f1_clean$driverRef == "max_verstappen"])
population_mean <- mean(f1_clean$positionOrder)
obs_diff <- population_mean - verstappen_mean # Positive means Verstappen is faster (lower position)

# Permutation loop
set.seed(123)
n_perms <- 2000
perm_diffs <- numeric(n_perms)

for(i in 1:n_perms) {
  # Shuffle the driver labels
  shuffled_drivers <- sample(f1_clean$driverRef)
  
  # Calculate mean for "fake" Hamiltons in this shuffled set
  perm_verstappebn_mean <- mean(f1_clean$positionOrder[shuffled_drivers == "max_verstappen"])
  perm_diffs[i] <- population_mean - perm_verstappebn_mean
}

# Calculate p-value
p_val <- mean(perm_diffs >= obs_diff)
print(paste("Observed Diff:", round(obs_diff, 2)))
print(paste("P-value:", p_val))

```


##### plot
```{r}
par(mfrow=c(2,2)) 

# 1. Histogram: Distribution of Fastest Lap Speeds
hist(f1_clean$fastestLapSpeed, 
     main = "Histogram of Fastest Lap Speeds", 
     xlab = "Speed (km/h)", col = "skyblue", border = "white")

# 2. Boxplot: Starting Grid Position vs. Did they Score Points?
boxplot(grid ~ is_points, data = f1_clean,
        main = "Grid Position vs. Points Finish",
        xlab = "Scored Points (0=No, 1=Yes)", ylab = "Grid Position",
        col = c("lightcoral", "skyblue"))


boxplot(grid ~ driverRef , data = f1_clean,
        main = "Grid Position vs. Points Finish",
        xlab = "Scored Points (0=No, 1=Yes)", ylab = "Grid Position",
        col = c("lightcoral", "skyblue"))

# 3. Scatterplot: Grid Position vs. Finishing Position
plot(f1_clean$grid, f1_clean$positionOrder,
     main = "Scatterplot: Grid vs. Finish",
     xlab = "Starting Grid Position", ylab = "Finishing Position",
     pch = 19, col = rgb(0,0,0,0.2)) 

# 4. Normal Quantile Plot (QQ Plot)
# Checking if Lap Speeds are normally distributed
qqnorm(f1_clean$fastestLapSpeed, main = "QQ Plot: Lap Speeds")
qqline(f1_clean$fastestLapSpeed, col = "red")

```

##### Driver Quali vs Race Result Analysis
Question: how strong does qualifying performance (grid) predict race performance (positionOrder) for specific drivers, and overall? 

Driver 1: Lewis Hamilton
```{r}

driver_data_hamilton <- f1_clean %>% filter(driverRef == "hamilton")
correlation <- cor(driver_data_hamilton$grid, driver_data_hamilton$positionOrder, use = "complete.obs")

ggplot(driver_data_hamilton, aes(x = grid, y = positionOrder)) + 
  geom_point(color = "lightblue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "orange", se = FALSE) +
  labs(
    title = paste("Qualifying vs. Race Result:", toupper("Lewis Hamilton")),
    subtitle = paste("Correlation:", round(correlation, 2)),
    x = "Starting Grid Position", 
    y = "Finishing Position") +
  theme_bw()
  
```


##### fastest average speed by circuit
Question: which circuits are the fastest? speed distribution

Fastest lap speed across all tracks
```{r}


# fastest lap speed comparison across all tracks
f1_clean %>%
  ggplot(aes(x = race_name, y = fastestLapSpeed, fill = race_name)) +
  geom_boxplot(outlier.size = 0.5, alpha = 0.8) +
  labs(title = "Fastest Lap Speed Comparison across tracks",
       x = NULL,
       y = "Fastest Lap Speed (km/h)") +
  coord_flip() +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.y = element_text(size = 6))
```

```{r}
# top 10 activity teams
top_teams <- f1_clean %>%
  count(constructorRef, sort = TRUE) %>%
  top_n(10) %>%
  pull(constructorRef)

# par(mar = c(8,4,4,2))
boxplot(positionOrder ~ constructorRef, 
        data = f1_clean %>% filter(constructorRef %in% top_10_teams),
        main = "Finishing Position by Team (2010-2024)",
        ylab = "Finishing Position",
        col = "lightblue",
        las = 2) 

f1_clean %>%
  filter(constructorRef %in% top_teams) %>% 
  mutate(constructorRef = reorder(constructorRef, positionOrder, median)) %>% 
  ggplot(aes(x = constructorRef, y = positionOrder, fill = constructorRef)) +
  geom_boxplot(ooutlier.alpha = 0.3) +
  labs(title = "Finishing Position Distribution by Team (2010-2024)",
       subtitle = "Ordered by Median Finishing Position (Lower is Better)",
       x = "Team",
       y = "Finishing Position") +
  theme_bw() +
  coord_flip() 

```

##### [explorative] f1dataR package 
```{r}
# install.packages("f1dataR")
# library(f1dataR)

if (!require("remotes")) install.packages("remotes")
remotes::install_github("SCasanova/f1dataR")
library(f1dataR)
```

Telemetry plotting
```{r}
get_driver_name()

f1_laps <- load_laps(season = 2025, round = 15)
f1_driver_telemetry <- load_driver_telemetry(season = "current", round = "last", session = "R", driver = "all", laps = "all")

ham <- load_driver_telemetry(
    season = 2023,
    round = "Bahrain",
    session = "Q",
    driver = "HAM",
    laps = "fastest"
  )
ham <- load_driver_telemetry(2022, 1, "Q", driver = "HAM", laps = "fastest")
per <- load_driver_telemetry(2022, 1, "Q", driver = "PER", laps = "fastest")

telem <- bind_rows(lec, ham, per) %>%
  select(rpm, speed, n_gear, throttle, brake, drs, distance, time, driver_code) %>%
  mutate(drs = ifelse(drs == 12, 1, 0))

drivercolours <- c(
  "LEC" = get_driver_color("LEC", 2022, 1),
  "HAM" = get_driver_color("HAM", 2022, 1),
  "PER" = get_driver_color("PER", 2022, 1)
)
```


